---
title: "Bayside census analysis"
runningheader: "Bayside census analysis" # only for pdf output
subtitle: "Working note" # only for html output
author: "Dr James Reynolds"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(dplyr)
library(ggstatsplot)
library(gtsummary)
library(janitor)
library(knitr)
library(kableExtra)
library(readr)
library(tidyverse)
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

# Introduction
The Public Transport Research Group (PTRG) is reviewing census data for 2011, 2016 and 2021 for Greater Melbourne and the City of Bayside. Journey to work travel, mode and total cars per person are the particular areas of interest.  

This working note details the analysis and results. 

# Methodology
Australian Bureau of Statistics (ABS) census data was obtained using the TableBuilder platform. Details are as follows


## Population and population over 18

Population of driving age was extracted from the census results that are enumerated by place of usual residence. Data was loaded and cleaned for 2011, 2016 and 2021.


```{r population_load_data, message=FALSE, warning=FALSE, echo=FALSE}

gccsa_ur_agep_2021 <- read_csv("data/gccsa_ur_agep_2021.csv", 
    skip = 8)
#remove unneeded rows and columns and clean names
gccsa_ur_agep_2021 <- gccsa_ur_agep_2021[-1,]
gccsa_ur_agep_2021 <- gccsa_ur_agep_2021[1:34,1:118]
gccsa_ur_agep_2021 <- gccsa_ur_agep_2021 %>% clean_names()

gccsa_ur_agep_2016 <- read_csv("data/gccsa_ur_agep_2016.csv", 
    skip = 8)
#remove unneeded rows and columns and clean names
gccsa_ur_agep_2016 <- gccsa_ur_agep_2016[-1,]
gccsa_ur_agep_2016 <- gccsa_ur_agep_2016[1:34,1:118]
gccsa_ur_agep_2016 <- gccsa_ur_agep_2016 %>% clean_names()

gccsa_ur_agep_2011 <- read_csv("data/gccsa_ur_agep_2011.csv", 
    skip = 8)
#remove unneeded rows and columns and clean names
gccsa_ur_agep_2011 <- gccsa_ur_agep_2011[-1,]
gccsa_ur_agep_2011 <- gccsa_ur_agep_2011[1:34,1:118]
gccsa_ur_agep_2011 <- gccsa_ur_agep_2011 %>% clean_names()

lga_ur_agep_2021 <- read_csv("data/lga_ur_agep_2021.csv", 
    skip = 8)
#remove unneeded rows and columns and clean names
lga_ur_agep_2021 <- lga_ur_agep_2021[-1,]
lga_ur_agep_2021 <- lga_ur_agep_2021[1:565,1:118]
lga_ur_agep_2021 <- lga_ur_agep_2021 %>% clean_names()

lga_ur_agep_2016 <- read_csv("data/lga_ur_agep_2016.csv", 
    skip = 8)
#remove unneeded rows and columns and clean names
lga_ur_agep_2016 <- lga_ur_agep_2016[-1,]
lga_ur_agep_2016 <- lga_ur_agep_2016[1:563,1:118]
lga_ur_agep_2016 <- lga_ur_agep_2016 %>% clean_names()

lga_ur_agep_2011 <- read_csv("data/lga_ur_agep_2011.csv", 
    skip = 8)
#remove unneeded rows and columns and clean names
lga_ur_agep_2011 <- lga_ur_agep_2011[-1,]
lga_ur_agep_2011 <- lga_ur_agep_2011[1:577,1:118]
lga_ur_agep_2011 <- lga_ur_agep_2011 %>% clean_names()

#fix column names
colnames(gccsa_ur_agep_2021)[colnames(gccsa_ur_agep_2021) == "agep_age"] <- "area_of_interest"
colnames(gccsa_ur_agep_2016)[colnames(gccsa_ur_agep_2016) == "agep_age"] <- "area_of_interest"
colnames(gccsa_ur_agep_2011) <- colnames(gccsa_ur_agep_2016)
colnames(lga_ur_agep_2021)[colnames(lga_ur_agep_2021) == "agep_age"] <- "area_of_interest"
colnames(lga_ur_agep_2016)[colnames(lga_ur_agep_2016) == "agep_age"] <- "area_of_interest"
colnames(lga_ur_agep_2011) <- colnames(lga_ur_agep_2016)

#fix variable types
gccsa_ur_agep_2021 <- gccsa_ur_agep_2021 %>% type.convert(as.is = T)
gccsa_ur_agep_2016 <- gccsa_ur_agep_2016 %>% type.convert(as.is = T)
gccsa_ur_agep_2011 <- gccsa_ur_agep_2011 %>% type.convert(as.is = T)
lga_ur_agep_2021 <- lga_ur_agep_2021 %>% type.convert(as.is = T)
lga_ur_agep_2016 <- lga_ur_agep_2016 %>% type.convert(as.is = T)
lga_ur_agep_2011 <- lga_ur_agep_2011 %>% type.convert(as.is = T)



```

## Cars per dwelling
The number of cars is enumerated in the census for each dwelling^[See https://www.abs.gov.au/census/guide-census-data/census-dictionary/2021/variables-topic/transport/number-motor-vehicles-vehd]. Data was loaded and cleaned for 2011, 2016 and 2021.

```{r cars_per_dwelling_load_data, message=FALSE, warning=FALSE, echo=FALSE}

gccsa_dw_vehd_2021 <- read_csv("data/gccsa_dw_vehd_2021.csv", 
    skip = 8)
#remove unneeded rows and columns and clean names
gccsa_dw_vehd_2021 <- gccsa_dw_vehd_2021[-1,]
gccsa_dw_vehd_2021 <- gccsa_dw_vehd_2021[1:25,1:34]
gccsa_dw_vehd_2021 <- gccsa_dw_vehd_2021 %>% clean_names()

gccsa_dw_vehd_2016 <- read_csv("data/gccsa_dw_vehd_2016.csv", 
    skip = 8)
#remove unneeded rows and columns and clean names
gccsa_dw_vehd_2016 <- gccsa_dw_vehd_2016[-1,]
gccsa_dw_vehd_2016 <- gccsa_dw_vehd_2016[1:25,1:34]
gccsa_dw_vehd_2016 <- gccsa_dw_vehd_2016 %>% clean_names()

gccsa_dw_vehd_2011 <- read_csv("data/gccsa_dw_vehd_2011.csv", 
    skip = 8)
#remove unneeded rows and columns and clean names
gccsa_dw_vehd_2011 <- gccsa_dw_vehd_2011[-1,]
gccsa_dw_vehd_2011 <- gccsa_dw_vehd_2011[1:25,1:34]
gccsa_dw_vehd_2011 <- gccsa_dw_vehd_2011 %>% clean_names()


lga_dw_vehd_2021 <- read_csv("data/lga_dw_vehd_2021.csv", 
    skip = 8)
#remove unneeded rows and columns and clean names
lga_dw_vehd_2021 <- lga_dw_vehd_2021[-1,]
lga_dw_vehd_2021 <- lga_dw_vehd_2021[1:556,1:34]
lga_dw_vehd_2021 <- lga_dw_vehd_2021 %>% clean_names()


lga_dw_vehd_2016 <- read_csv("data/lga_dw_vehd_2016.csv", 
    skip = 8)
#remove unneeded rows and columns and clean names
lga_dw_vehd_2016 <- lga_dw_vehd_2016[-1,]
lga_dw_vehd_2016 <- lga_dw_vehd_2016[1:554,1:34]
lga_dw_vehd_2016 <- lga_dw_vehd_2016 %>% clean_names()


lga_dw_vehd_2011 <- read_csv("data/lga_dw_vehd_2011.csv", 
    skip = 8)
#remove unneeded rows and columns and clean names
lga_dw_vehd_2011 <- lga_dw_vehd_2011[-1,]
lga_dw_vehd_2011 <- lga_dw_vehd_2011[1:568,1:34]
lga_dw_vehd_2011 <- lga_dw_vehd_2011 %>% clean_names()




```



# Results


## Population and population 18 and over

The following tables and figures show the population that is 18 years and older and the total populations in both the City of Bayside and the rest of Greater Melbourne. 

```{r population_of_driving_age_summarise_data, message=FALSE, warning=FALSE, echo=FALSE, fig.caption = "Total population 18 and older"}

driving_age_summary_table <- gccsa_ur_agep_2021 %>%
  filter(area_of_interest == "Greater Melbourne")
driving_age_summary_table <- driving_age_summary_table %>% 
  add_row(gccsa_ur_agep_2016 %>%
  filter(area_of_interest == "Greater Melbourne"))
driving_age_summary_table <- driving_age_summary_table %>% 
  add_row(gccsa_ur_agep_2011 %>%
  filter(area_of_interest == "Greater Melbourne"))
driving_age_summary_table <- driving_age_summary_table %>% 
  add_row(lga_ur_agep_2021 %>%
  filter(area_of_interest == "Bayside (Vic.)"))
driving_age_summary_table <- driving_age_summary_table %>% 
  add_row(lga_ur_agep_2016 %>%
  filter(area_of_interest == "Bayside (C)"))
driving_age_summary_table <- driving_age_summary_table %>% 
  add_row(lga_ur_agep_2011 %>%
  filter(area_of_interest == "Bayside (C)"))

driving_age_summary_table$Year <- c(2021, 2016, 2011, 2021, 2016, 2011)

#replace first column with consistent place names
driving_age_summary_table[["area_of_interest"]] <- c("Greater Melbourne", "Greater Melbourne", "Greater Melbourne", "Bayside", "Bayside", "Bayside")

driving_age_summary_table$driving_age_population <- driving_age_summary_table[,20:117] %>% rowSums()


total_population_summary_table <- driving_age_summary_table %>% 
  select(area_of_interest, Year, total)


driving_age_summary_table <- driving_age_summary_table %>% 
  select(area_of_interest, Year, driving_age_population)

driving_age_summary_table_wider <- driving_age_summary_table %>% 
  select(area_of_interest, Year, driving_age_population) %>% 
  pivot_wider(
  names_from = area_of_interest, 
  values_from = driving_age_population
)


driving_age_summary_table_wider$`Rest of Greater Melbourne` <- driving_age_summary_table_wider$`Greater Melbourne` - driving_age_summary_table_wider$Bayside
 
driving_age_summary_table_wider <- driving_age_summary_table_wider %>% 
  select(Year, Bayside, `Rest of Greater Melbourne`)

driving_age_summary_table_wider$Year <- driving_age_summary_table_wider$Year %>%   factor(levels = c("2011", "2016", "2021")) 

driving_age_summary_table_wider <- driving_age_summary_table_wider %>% 
  arrange(Year)

driving_age_summary_table_wider %>%   
  adorn_totals(where = "col") %>% 
  adorn_percentages(denominator = "row") %>% 
  adorn_pct_formatting() %>% 
  adorn_ns() %>%
  kable(caption = "Total population 18 and older", 
        align = 'lrr', 
        format.args = list(big.mark = ","))

 ggbarstats(driving_age_summary_table_wider %>% pivot_longer(
    cols = 2:3,
    names_to = "area_of_interest", 
    values_to = "Total population 18 and older"),
  y = Year, 
  x = area_of_interest, 
  counts = "Total population 18 and older")

```
 



```{r population_total_summarise_data, message=FALSE, warning=FALSE, echo=FALSE, fig.caption = "Total population"}

 
#### Total Popluation 
 
total_population_summary_table_wider <- total_population_summary_table %>% 
  pivot_wider(
  names_from = area_of_interest, 
  values_from = total
)


total_population_summary_table_wider$`Rest of Greater Melbourne` <- total_population_summary_table_wider$`Greater Melbourne` - total_population_summary_table_wider$Bayside
 
total_population_summary_table_wider <- total_population_summary_table_wider %>% 
  select(Year, Bayside, `Rest of Greater Melbourne`)

total_population_summary_table_wider$Year <- total_population_summary_table_wider$Year %>%   factor(levels = c("2011", "2016", "2021")) 

total_population_summary_table_wider <- total_population_summary_table_wider %>% 
  arrange(Year)

total_population_summary_table_wider %>%   
  adorn_totals(where = "col") %>% 
  adorn_percentages(denominator = "row") %>% 
  adorn_pct_formatting() %>% 
  adorn_ns() %>%
  kable(caption = "Total population", 
        align = 'lrr', 
        format.args = list(big.mark = ","))

 ggbarstats(total_population_summary_table_wider %>% pivot_longer(
    cols = 2:3,
    names_to = "area_of_interest", 
    values_to = "Total population"),
  y = Year, 
  x = area_of_interest, 
  counts = "Total population")
```

The following tables and figures compare the share of the population that is over or under the age of 18. 

```{r population_comparisons, message=FALSE, warning=FALSE, echo=FALSE, fig.caption = "Population share: younger than 18 compared to 18 and over", fig.fullwidth = TRUE, fig.height = 10}

 population_comparison_table <- driving_age_summary_table_wider %>%
   adorn_totals(where = "col")
 population_comparison_table$total_population_bayside <- total_population_summary_table_wider$Bayside
  population_comparison_table$total_population_rest_of_greater_melbourne <- total_population_summary_table_wider$`Rest of Greater Melbourne`
 population_comparison_table$total_population_greater_melbourne <- 
   total_population_summary_table_wider$Bayside +
   total_population_summary_table_wider$`Rest of Greater Melbourne`
   
names(population_comparison_table) <- c(
  "Year", 
  "x18plus_bayside", 
  "x18plus_rest_of_melbourne", 
  "x18plus_greater_melbourne", 
  "total_population_bayside", 
  "total_population_rest_of_melbourne", 
  "total_population_greater_melbourne") 

population_comparison_table$x18plus_share_bayside <- population_comparison_table$x18plus_bayside / population_comparison_table$total_population_bayside

population_comparison_table$x18plus_share_rest_of_melbourne <- population_comparison_table$x18plus_rest_of_melbourne / population_comparison_table$total_population_rest_of_melbourne

population_comparison_table$x18plus_share_greater_melbourne <- population_comparison_table$x18plus_greater_melbourne / population_comparison_table$total_population_greater_melbourne

# plot giant comparison table
kable(population_comparison_table %>% 
        adorn_pct_formatting(,,,8:10), 
      caption = "City of Bayside and Rest of Greater Melbourne: population 18 and over, total, and share 18 and over",
      align = "lrrrrrrrrrr",
      format.args = list(big.mark = ",", scientific = FALSE),
      col.names = c("Year", 
                   "City of Bayside", "Rest of Greater Melbourne", "Total",
                   "City of Bayside", "Rest of Greater Melbourne", "Total",
                   "City of Bayside", "Rest of Greater Melbourne", "Total")
      ) %>% 
  column_spec(column = 8:10, ) %>%
  add_header_above(c("Census" =1, "18 years and older" = 3, "Total population" = 3, "Share 18 years and older" = 3))


# compares shares
#share_18_and_old <- population_comparison_table %>% 
#  select(Year, 
#  x18plus_bayside, 
#  x18plus_rest_of_melbourne, 
#  total_population_bayside, 
#  total_population_rest_of_melbourne)

#share_18_and_old$x17minus_bayside <- share_18_and_old$total_population_bayside - share_18_and_old$x18plus_bayside
#share_18_and_old$x17minus_rest_of_melbourne <- share_18_and_old$total_population_rest_of_melbourne - share_18_and_old$x18plus_rest_of_melbourne

#share_18_and_old <- share_18_and_old %>% select(
#  Year, 
#  x17minus_bayside, 
#  x18plus_bayside,
#  x17minus_rest_of_melbourne,
#  x18plus_rest_of_melbourne
#)

#share_18_and_old

#kable(share_18_and_old %>% 
#        untabyl() %>%
#        adorn_totals(where = "col") %>%
#        adorn_percentages() %>% 
#        adorn_pct_formatting() %>% 
#        adorn_ns, 
#      caption = "City of Bayside and Rest of Greater Melbourne: population younger and older than 18",
#      align = "lrrrrr",
#      col.names = c("Year", 
#                   "under 18", "over 18", 
#                   "under 18", "over 18",
#                   "Total")) %>%
#        add_header_above(c("Census" = 1, "City of Bayside" = 2, "Rest of Melbourne" = 2, "Grand" = 1))



# compares shares
share_18_and_old_bayside <- population_comparison_table %>% 
  select(Year, 
  x18plus_bayside, 
  total_population_bayside)

share_18_and_old_bayside$x17_minus <- share_18_and_old_bayside$total_population_bayside - share_18_and_old_bayside$x18plus_bayside
share_18_and_old_bayside$area_of_interest <- "City of Bayside"
names(share_18_and_old_bayside) <- c("Year", "x18_plus", "total", "x17_minus", "area_of_interest")


share_18_and_old_rest_of_melbourne <- population_comparison_table %>% 
  select(Year, 
  x18plus_rest_of_melbourne, 
  total_population_rest_of_melbourne)
share_18_and_old_rest_of_melbourne$x17_minus <- share_18_and_old_rest_of_melbourne$total_population_rest_of_melbourne - share_18_and_old_rest_of_melbourne$x18plus_rest_of_melbourne
share_18_and_old_rest_of_melbourne$area_of_interest <- "Rest of Melbourne"
names(share_18_and_old_rest_of_melbourne) <- c("Year", "x18_plus", "total", "x17_minus", "area_of_interest")

share_18_and_old <- add_row(share_18_and_old_bayside %>% 
                              select(Year, area_of_interest, x17_minus, x18_plus), 
                            share_18_and_old_rest_of_melbourne %>% 
                              select(Year, area_of_interest, x17_minus, x18_plus))

share_18_and_old <- share_18_and_old %>% arrange(Year)


kable(share_18_and_old %>% 
        untabyl() %>%
        adorn_totals(where = "col") %>%
        adorn_percentages() %>% 
        adorn_pct_formatting() %>% 
        adorn_ns, 
      caption = "City of Bayside and Rest of Greater Melbourne: population younger and older than 18",
      align = "llrrr",
      col.names = c("Year", "Location", "under 18", "over 18", "Total"))

share_18_and_old_longer <- share_18_and_old %>% 
  pivot_longer(cols = 3:4,
               names_to = "group", 
               values_to = "population")

grouped_ggbarstats(data = share_18_and_old_longer, 
                   x = group, 
                   y = area_of_interest, 
                   counts = population, 
                   grouping.var = Year, 
                   plotgrid.args = list(nrow = 3), 
                   xlab = "Age group", 
                   ylab = "Area")
                   

```

## Cars per dwelling

```{r cars_per_dwelling_summarise_data, message=FALSE, warning=FALSE, echo=FALSE}

cars_per_dwelling_summary_table <- gccsa_dw_vehd_2021 %>%
  filter(vehd_number_of_motor_vehicles == "Greater Melbourne")

cars_per_dwelling_summary_table <- cars_per_dwelling_summary_table %>% 
  add_row(gccsa_dw_vehd_2016 %>%
  filter(vehd_number_of_motor_vehicles == "Greater Melbourne"))

cars_per_dwelling_summary_table <- cars_per_dwelling_summary_table %>% 
  add_row(gccsa_dw_vehd_2011 %>%
  filter(vehd_number_of_motor_vehicles == "Greater Melbourne"))

cars_per_dwelling_summary_table <- cars_per_dwelling_summary_table %>% 
  add_row(lga_dw_vehd_2021 %>%
  filter(vehd_number_of_motor_vehicles == "Bayside (Vic.)"))

cars_per_dwelling_summary_table <- cars_per_dwelling_summary_table %>% 
  add_row(lga_dw_vehd_2016 %>%
  filter(vehd_number_of_motor_vehicles == "Bayside (C)"))

cars_per_dwelling_summary_table <- cars_per_dwelling_summary_table %>% 
  add_row(lga_dw_vehd_2011 %>%
  filter(vehd_number_of_motor_vehicles == "Bayside (C)"))

cars_per_dwelling_summary_table$Year <- c(2021, 2016, 2011, 2021, 2016, 2011)

#replace first column with consistent place names
area_of_interest <- c("Greater Melbourne", "Greater Melbourne", "Greater Melbourne", "Bayside", "Bayside", "Bayside")
cars_per_dwelling_summary_table[["vehd_number_of_motor_vehicles"]] <- area_of_interest
colnames(cars_per_dwelling_summary_table)[colnames(cars_per_dwelling_summary_table) == "vehd_number_of_motor_vehicles"] <- "area_of_interest"

cars_per_dwelling_summary_table$total_cars <- 0

cars_per_dwelling_summary_table$total_cars <- cars_per_dwelling_summary_table$x1_motor_vehicle + 
  cars_per_dwelling_summary_table$x2_motor_vehicles * 2 + 
  cars_per_dwelling_summary_table$x3_motor_vehicles * 3 + 
  cars_per_dwelling_summary_table$x4_motor_vehicles * 4 + 
  cars_per_dwelling_summary_table$x5_motor_vehicles * 5 + 
  cars_per_dwelling_summary_table$x6_motor_vehicles * 6 + 
  cars_per_dwelling_summary_table$x7_motor_vehicles * 7 + 
  cars_per_dwelling_summary_table$x8_motor_vehicles * 8 + 
  cars_per_dwelling_summary_table$x9_motor_vehicles * 9 + 
  cars_per_dwelling_summary_table$x10_motor_vehicles * 10 + 
  cars_per_dwelling_summary_table$x11_motor_vehicles * 11 + 
  cars_per_dwelling_summary_table$x12_motor_vehicles * 12 + 
  cars_per_dwelling_summary_table$x13_motor_vehicles * 13 + 
  cars_per_dwelling_summary_table$x14_motor_vehicles * 14 + 
  cars_per_dwelling_summary_table$x15_motor_vehicles * 15 + 
  cars_per_dwelling_summary_table$x16_motor_vehicles * 16 + 
  cars_per_dwelling_summary_table$x17_motor_vehicles * 17 + 
  cars_per_dwelling_summary_table$x18_motor_vehicles * 18 + 
  cars_per_dwelling_summary_table$x19_motor_vehicles * 19 + 
  cars_per_dwelling_summary_table$x20_motor_vehicles * 20 + 
  cars_per_dwelling_summary_table$x21_motor_vehicles * 21 + 
  cars_per_dwelling_summary_table$x22_motor_vehicles * 22 + 
  cars_per_dwelling_summary_table$x23_motor_vehicles * 23 + 
  cars_per_dwelling_summary_table$x24_motor_vehicles * 24 + 
  cars_per_dwelling_summary_table$x25_motor_vehicles * 25 + 
  cars_per_dwelling_summary_table$x26_motor_vehicles * 26 + 
  cars_per_dwelling_summary_table$x27_motor_vehicles * 27 + 
  cars_per_dwelling_summary_table$x28_motor_vehicles * 28 + 
  cars_per_dwelling_summary_table$x29_motor_vehicles * 29 + 
  cars_per_dwelling_summary_table$x30_or_more_motor_vehicles * 30 

summary_table <- cars_per_dwelling_summary_table %>% 
  select(area_of_interest, Year, total_cars)

total_cars_summary_table <- cars_per_dwelling_summary_table %>% 
  select(area_of_interest, Year, total_cars) %>% 
  pivot_wider(
  names_from = area_of_interest, 
  values_from = total_cars
)


total_cars_summary_table$`Rest of Greater Melbourne` <- total_cars_summary_table$`Greater Melbourne` - total_cars_summary_table$Bayside
 
total_cars_summary_table <- total_cars_summary_table %>% 
  select(Year, Bayside, `Rest of Greater Melbourne`)

total_cars_summary_table$Year <- total_cars_summary_table$Year %>% 
  factor(levels = c("2011", "2016", "2021")) 

total_cars_summary_table <- total_cars_summary_table %>% 
  arrange(Year)

total_cars_summary_table %>%   
  adorn_totals(where = "col") %>% 
  adorn_percentages(denominator = "row") %>% 
  adorn_pct_formatting() %>% 
  adorn_ns() %>%
  kable(caption = "Total number of cars by census", 
        align = 'lrr', 
        format.args = list(big.mark = ","))

# ggbarstats(total_cars_summary_table %>% pivot_longer(
#    cols = 2:3,
#    names_to = "area_of_interest", 
#    values_to = "Total cars"),
#  y = Year, 
#  x = area_of_interest, 
#  counts = "Total cars")

```



# Discussion and Conclusions

# References

```{r bib, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'skeleton.bib')
```
